---
title: "A Socio-Economic Model of Residential Segregation, Neighborhood Change and Housing Inequality"
subtitle: "Empirical Analysis: Individual-Level Hypotheses (ACS PUMS)"
author: "Malte Grönemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo    = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      error   = FALSE,
                      out.width = "100%")

# Packages
library(nanoparquet)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggcorrplot)
library(fixest)
library(modelsummary)
library(latex2exp)
library(kableExtra)

# Plotting
cb_palette <- c("#0072B2", "#D55E00", "#009E73", "#CC79A7", "#E69F00", "#56B4E9")
lines <- c("solid", "twodash", "longdash", "dotdash", "dotted")
shapes <- c(0, 1, 2, 5, 6)

# File Paths
# TODO for replication: change paths
# local paths for development
#data_dir <- "~/Projects/SEHTestACS/data/"
#results_dir <- "../results/"
# paths at HELIX HPC
data_dir <- "./data/"
results_dir <- "./results/"

# Data
pums <- read_parquet(paste0(data_dir, "pums_analysis.parquet")) |>
  mutate(YEAR = as.integer(YEAR),
         # race_eth: nhwhite as reference category
         race_eth = factor(race_eth, levels = c("nhwhite", "nhblack", "hispanic", "nhother"))
         )

#acs <- read_parquet(paste0(data_dir, "acs_top100.parquet"))

cbsa <- read_parquet(paste0(data_dir, "cbsa_vars.parquet"))
```


# Data and Operationalisation

## Housing Quality

The ACS PUMS does not collect housing quality directly. We use market-based proxies that capture both unit characteristics and neighbourhood desirability although the original hypotheses on housing quality were based on characteristics of the unit alone. Please see the analysis for utility and rent for those. In my model the relationship of income and rent is not moderated by segregation because rent is a perfect reflection of desirability in my model and the richest households will sort into the best units no matter why they are desirable or how segregated the city is. But WHY OK HERE?

- **Renters:** gross monthly rent in 1999 USD (`rentgrs_real`)
- **Owners:** self-reported home value in 1999 USD (`valueh_real`)

To remove between-metro and between-year price level differences, both measures are z-standardised within YEAR × CBSA × tenure group. The resulting `quality` variable is a within-context quality rank (mean 0, SD 1 in each group).

These measures are definitely imperfect measures of housing quality. Self-reported home values are likely biased by their owners' perception (LIT ON SUBJECTIVE PRICES?) and the rent a household pays might not be a perfect representation of market value when they have an older contract and/or the unit is under rent control. I nonetheless think that on average, higher values on these measures coincides with a higher quality/desirability of the unit. The measurement error likely attenuates the estimates though.

Household income is equivalised using the square root household size scale (`hhincome_real / sqrt(NUMPREC)`) and log-transformed to account for the well-documented diminishing returns in the income–housing relationship. Households with zero or negative income are bottom-coded at 1 (1999 USD), giving a floor of log(1) = 0. Income is top-coded at the within-CBSA × year p99 to reduce the influence of extreme values.

**Robustness check on rent control:** In cities with rent control (notably New York, San Francisco, and Los Angeles), long-term tenants may pay substantially below market rent, detaching rent from current housing quality. A robustness version of the quality variable (`quality_rc`) standardises renter quality using only households who moved into their unit within the past two years (`MOVEDIN >= YEAR - 2`), where rents more closely reflect current market conditions.

## Sample Coverage

```{r coverage}
coverage <- pums |>
  mutate(tenure = if_else(renter == 1L, "Renters", "Owners")) |>
  count(CBSA_name, YEAR, tenure) |>
  arrange(CBSA_name, YEAR, tenure)

summary(coverage$n)
```

The minimum group size across all CBSA × year × tenure cells is `r min(coverage$n)` households, sufficient for reliable within-group standardisation.


# Rent/Housing Segregation > Income Segregation

> Rent and Housing Quality are more unequally distributed in space within cities than household income.

problems: no uniform size of the housing unit, hh income means different things for hh of different size/composition, only gross rent and hh income in ACS. However, unit size is endogenous to the developers' decision as well. And to pay absolute rets, hh income matters not equiv income.

Spatial unevenness is measured by the rank-order information theory index H^R (Reardon & Bischoff 2011) for the continuous variables income, gross rent, and home value, and by the binary Theil H for the categorical variable tenure. Both indices are bounded [0,1] and scale-independent: H^R is robust to bracket boundary shifts and the absolute level of the variable, so cross-city and over-time comparisons are valid without deflating to constant dollars. The two indices are conceptually consistent — H^R for a binary variable reduces exactly to binary H — but not identical: H^R summarises segregation across the full distributional gradient by integrating binary H values across all possible cutpoints, whereas H for tenure captures a single renter/owner split. Comparisons between H^R and H are therefore meaningful as measures of overall spatial unevenness but should not be interpreted as the same underlying quantity. A further difference in universe should be noted: H_inc is computed over all households, whereas H_rent covers only cash-rent-paying renters (B25063) and H_val only owner-occupiers (B25075). In principle this could inflate the housing indices if renters or owners are already more spatially concentrated as subpopulations. In practice, however, H^R is distribution-invariant — it captures how unevenly the *within-universe* distribution of rents or values is spread across tracts, not how the universe itself is placed — so the comparison remains meaningful as a test of whether each housing dimension is more spatially unequal than income within its respective population.

## Description of Segregation Indices

Over all metros and periods, home value is most unevenly distributed of the continuous variables ($H^R$ = `r round(mean(cbsa$H_val, na.rm = T), 3)`), followed by rent ($H^R$ = `r round(mean(cbsa$H_rent, na.rm = T), 3)`) and income ($H^R$ = `r round(mean(cbsa$H_inc, na.rm = T), 3)`; $H(0.5)$ = `r round(mean(cbsa$H_inc_p50, na.rm = T), 3)`). Home value and rents are really substantially uneven among tracts. Tenure type is also substantively unequal between census tracts with $H$ = `r round(mean(cbsa$H_tenure, na.rm = T), 3)`. Racial/ethnic segregation is very prevalent in the US and amounts to $H$ = `r round(mean(cbsa$H_bw, na.rm = T), 3)` when measured between Whites and Blacks only and to $H$ = `r round(mean(cbsa$H_race_multi, na.rm = T), 3)` for the multi-group Theil index on average. Vacancies are less unevenly placed with  $H$ = `r round(mean(cbsa$H_vac, na.rm = T), 3)`.

```{r}
cbsa |>
  select(H_inc, H_rent, H_val, H_tenure, H_bw, H_race_multi) |>
  cor(use = "complete.obs") |>
  ggcorrplot(type = "upper",
             show.legend = F, lab = T) +
  labs(title = "Correlation Matrix")
```
Cities with high income segregation tend to be segregated on the other dimensions as well, but the association is strongest with home values (0.77). Where income sorts households across tracts, property values track that sorting closely; tenure (0.53) and rents (0.47) follow at some distance.

The clearest asymmetry concerns rents. Home value segregation correlates 0.38–0.39 with the two racial measures, while rent segregation correlates 0.07 with both. Racially segregated cities are differentiated by what their housing is worth but not by what it rents for. Rent segregation is the weakest correlate throughout the matrix — 0.27 with tenure, 0.47 with income, both the lowest in their columns — so it may simply be a measure with little across-tract variation to distribute. Tenure segregation relates more closely to the economic dimensions (0.53 with income, 0.40 with value) than to the racial ones (0.25–0.31). At the metro level, tenure is primarily an economic axis.

The two racial measures correlate 0.90, which is partly definitional, since multigroup segregation contains the Black–White contrast as a component. More informative is that both relate almost identically to everything else — 0.42/0.47 with income, 0.38/0.39 with value, 0.31/0.25 with tenure, 0.07/0.07 with rent. Extending the measure to Hispanics and other groups changes nothing about how racial segregation maps onto the housing dimensions.


## Test of the Hypothesis


```{r}
cbsa |>
  select(CBSAFP, cbsa_name_short, year, H_inc, H_inc_p50, H_rent, H_val, H_tenure) |>
  mutate(cbsa_name_short = forcats::fct_reorder(cbsa_name_short, H_inc, .fun = mean, na.rm = TRUE)) |>
  pivot_longer(starts_with("H_"),
               names_to = "H_dimension",
               values_to = "H_index") |>
  mutate(H_dimension = case_when(H_dimension == "H_inc"     ~ "Household Income (H^R)",
                                 H_dimension == "H_inc_p50" ~ "Household Income H(0.5)",
                                 H_dimension == "H_rent"    ~ "Gross Rent",
                                 H_dimension == "H_val"     ~ "Home Value",
                                 H_dimension == "H_tenure"  ~ "Tenure Type")) |>
  ggplot() +
  aes(y = H_index,
      x = cbsa_name_short,
      colour = H_dimension,
      shape = H_dimension,
      group = H_dimension) +
  geom_point() +
  geom_line() +
  facet_grid(rows = vars(year)) +
  labs(x = NULL, y = NULL, colour = NULL, shape = NULL) +
  scale_color_manual(values = cb_palette,
                     labels = c("Household Income (H^R)"   = TeX("Household Income $H^R$"),
                                "Household Income H(0.5)"  = TeX("Household Income $H(0.5)$"),
                                "Gross Rent"               = TeX("Gross Rent $H^R$"),
                                "Home Value"               = TeX("Home Value $H^R$"),
                                "Tenure Type"              = TeX("Tenure Type $H$"))) +
  scale_fill_manual(values = cb_palette) +
  scale_shape_manual(values = shapes,
                     labels = c("Household Income (H^R)"   = TeX("Household Income $H^R$"),
                                "Household Income H(0.5)"  = TeX("Household Income $H(0.5)$"),
                                "Gross Rent"               = TeX("Gross Rent $H^R$"),
                                "Home Value"               = TeX("Home Value $H^R$"),
                                "Tenure Type"              = TeX("Tenure Type $H$"))) +
  scale_linetype_manual(values = lines) +
  theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
ggsave(paste0(results_dir, "seg_comparison.pdf"), height = 20, width = 35, units = "cm")
```

Rent is more unevenly distributed than income in `r 100 * sum(cbsa$H_rent > cbsa$H_inc, na.rm = T) / length(!is.na(cbsa$CBSAFP))` percent of all valid observations of the 100 cities and 4 periods. The estimated home value is spatially more uneven in `r 100 * sum(cbsa$H_val > cbsa$H_inc, na.rm = T) / length(!is.na(cbsa$CBSAFP))` percent of observations. Housing tenure (renter vs. owner-occupier) is more unevenly placed than household income ($H^R$) in `r 100 * sum(cbsa$H_tenure > cbsa$H_inc, na.rm = T) / length(!is.na(cbsa$CBSAFP))` percent of all valid cases. The like-for-like comparison using $H(0.5)$ — binary H at the median income cutpoint, directly comparable to the single-split tenure H — yields the same conclusion: tenure exceeds income segregation in `r 100 * sum(cbsa$H_tenure > cbsa$H_inc_p50, na.rm = T) / length(!is.na(cbsa$CBSAFP))` percent of observations.

To test whether these differences are statistically significant, I regress each pairwise difference (e.g. H_rent − H_inc) on an intercept with standard errors two-way clustered by CBSA and year. Two-way clustering accounts for within-metro correlation across waves and within-year correlation across metros. Because the outcome is already a difference between two indices, year fixed effects would only matter if national trends in the two indices diverge, which is absorbed by the clustering. The null hypothesis is that the mean difference is zero, and a significant intercept means we can reject that null. The test does not account for ACS sampling uncertainty within tracts — the H indices are treated as fixed — so it captures variation across metros and time, not measurement error in the indices themselves.

```{r seg_diff_tests}
modelsummary(
  list("Gross Rent − Income"        = feols(I(H_rent    - H_inc)    ~ 1, cluster = ~CBSAFP + year, data = cbsa),
       "Home Value − Income"        = feols(I(H_val     - H_inc)    ~ 1, cluster = ~CBSAFP + year, data = cbsa),
       "Tenure − Income (H^R)"      = feols(I(H_tenure  - H_inc)    ~ 1, cluster = ~CBSAFP + year, data = cbsa),
       "Tenure − Income H(0.5)"       = feols(I(H_tenure - H_inc_p50) ~ 1, cluster = ~CBSAFP + year, data = cbsa)),
  estimate    = "{estimate}{stars}",
  statistic   = c("({std.error})", "p {p.value}"),
  coef_rename = c("(Intercept)" = "Difference"),
  stars       = c("+" = 0.1, "*" = .05, "**" = .01, "***" = .001),
  gof_map     = c("nobs"),
  output      = "html"
)
```

## Robustness Check: Segregation within Race

In the US, the housing market is not only characterized by income segregation but also racial segregation. And the former (in combination with group differences in income) seems to be less a cause for the latter as often assumed. Instead, historical patterns of racially segregated neighborhoods reproduce itself (Krysan and Crowder 2017), effectively creating a segmented housing market. To see whether these relationships hold within race groups, I compare income and tenure segregation by three race groups: non-hispanic White, non-hispanic Black, and Hispanic/Latino. Home values and rents are not available as brackets separated by race (at least not in the publicly accessible ACS data) so that I cannot compare income segregation to rent/value segregation within race groups.


```{r}
cbsa |>
  select(H_inc, H_tenure, H_bw, H_race_multi, H_inc_nhw, H_tenure_nhw, H_inc_nhb, H_tenure_nhb, H_inc_his, H_tenure_his) |>
  cor(use = "complete.obs") |>
  ggcorrplot(type = "upper",
             show.legend = F, lab = T) +
  labs(title = "Correlation Matrix")
```
Tenure and income segregation move together far more tightly within racial groups than in the population as a whole: 0.53 overall, but 0.60 among Whites and 0.94–0.95 among Hispanics and Blacks. Within groups, sorting by tenure and sorting by income describe nearly the same partition.

Cities with high racial segregation are also cities in which the White population is internally stratified. Within-White tenure segregation correlates 0.80 with Black–White segregation and 0.41 with total tenure segregation, and within-White income segregation correlates 0.65 with total income segregation. For Blacks the associations run the other way. Within-Black income segregation correlates −0.29 with total income segregation and −0.52 with multigroup racial segregation; the tenure figures are −0.26 and −0.50. Cities that are highly segregated overall have comparatively economically mixed Black neighborhoods. These two patterns together are consistent with a white flight hypothesis: racial sorting and within-White tenure stratification produced jointly as White households distributed across a suburban ring differentiated by tenure and income, while Black households remained constrained across the income distribution, leaving middle-class and poor Black households in the same tracts.

Within-Hispanic segregation stands apart. Its two components correlate 0.94 with each other but only 0.07–0.31 with everything else, with signs that flip across the matrix — mildly positive with racial segregation, mildly negative with the within-White and within-Black measures. Cities that are segregated on the other dimensions are not systematically segregated within their Hispanic population.


```{r seg_race_plot}
cbsa |>
  select(CBSAFP, CBSA_name, year, H_inc,
         H_inc_nhw, H_inc_nhw_p50, H_inc_nhb, H_inc_nhb_p50, H_inc_his, H_inc_his_p50,
         H_tenure_nhw, H_tenure_nhb, H_tenure_his) |>
  mutate(CBSA_name = forcats::fct_reorder(CBSA_name, H_inc, .fun = mean, na.rm = TRUE)) |>
  select(-H_inc) |>
  pivot_longer(starts_with("H_"),
               names_to = "H_dimension",
               values_to = "H_index") |>
  mutate(
    race_group = case_when(grepl("nhw", H_dimension) ~ "NH-White",
                           grepl("nhb", H_dimension) ~ "NH-Black",
                           grepl("his", H_dimension) ~ "Hispanic"),
    H_type     = case_when(grepl("p50",    H_dimension) ~ "Income H(0.5)",
                           grepl("inc",    H_dimension) ~ "Income H^R",
                           grepl("tenure", H_dimension) ~ "Tenure")
  ) |> 
  ggplot() +
  aes(y = H_index,
      x = CBSA_name,
      colour = H_type,
      shape = H_type,
      group = H_type) +
  geom_point() +
  geom_line() +
  facet_grid(rows = vars(year), cols = vars(race_group)) +
  labs(x = "Metro Area", y = NULL, colour = NULL, shape = NULL) +
  scale_color_manual(values = cb_palette,
                     labels = c("Income H^R"    = TeX("Income $H^R$"),
                                "Income H(0.5)" = TeX("Income $H(0.5)$"),
                                "Tenure"        = TeX("Tenure $H$"))) +
  scale_fill_manual(values = cb_palette) +
  scale_shape_manual(values = shapes,
                     labels = c("Income H^R"    = TeX("Income $H^R$"),
                                "Income H(0.5)" = TeX("Income $H(0.5)$"),
                                "Tenure"        = TeX("Tenure $H$"))) +
  scale_linetype_manual(values = lines) +
  theme_bw() +
  theme(legend.position = "bottom",
        axis.text.x  = element_blank(),
        axis.ticks.x = element_blank())
```

The ordering of the figure is the same as above, based on overall income segregation. It shows that in general, tenure segregation exceeds income segregation within the three race groups. It also shows that both income and tenure segregation is much higher higher on average and much more varied between metro areas for Blacks and Hispanics compared to Whites. Averaged over all four ACS periods and all 100 metros, Hispanics have the highest within-group income segregation ($H^R_H$ = `r round(mean(cbsa$H_inc_his, na.rm = T), 3)`; $H(0.5)_H$ = `r round(mean(cbsa$H_inc_his_p50, na.rm = T), 3)`) with Blacks following closely ($H^R_B$ = `r round(mean(cbsa$H_inc_nhb, na.rm = T), 3)`; $H(0.5)_B$ = `r round(mean(cbsa$H_inc_nhb_p50, na.rm = T), 3)`). Whites in contrast only have a within-group income segregation index of $H^R_W$ = `r round(mean(cbsa$H_inc_nhw, na.rm = T), 3)` ($H(0.5)_W$ = `r round(mean(cbsa$H_inc_nhw_p50, na.rm = T), 3)`). Unevenness in tenure over census tracts is highest within Blacks ($H_B$ = `r round(mean(cbsa$H_tenure_nhb, na.rm = T), 3)`) but the difference to Hispanics is again small ($H_H$ = `r round(mean(cbsa$H_tenure_his, na.rm = T), 3)`). Tenure segregation of Whites is exactly half that of Blacks ($H_B$ = `r round(mean(cbsa$H_tenure_nhw, na.rm = T), 3)`). Within Hispanics, in `r 100 * sum(cbsa$H_tenure_his > cbsa$H_inc_his, na.rm = T) / length(!is.na(cbsa$CBSAFP))` percent of all valid observations over the four periods and 100 cities was tenure segregation higher than income segregation ($H^R$). For Blacks, that is the case in `r 100 * sum(cbsa$H_tenure_nhb > cbsa$H_inc_nhb, na.rm = T) / length(!is.na(cbsa$CBSAFP))` percent and for Whites in `r 100 * sum(cbsa$H_tenure_nhw > cbsa$H_inc_nhw, na.rm = T) / length(!is.na(cbsa$CBSAFP))` percent. The like-for-like $H(0.5)$ comparison strengthens this: tenure exceeds the within-group median-split income index in `r 100 * sum(cbsa$H_tenure_his > cbsa$H_inc_his_p50, na.rm = T) / length(!is.na(cbsa$CBSAFP))` percent of Hispanic, `r 100 * sum(cbsa$H_tenure_nhb > cbsa$H_inc_nhb_p50, na.rm = T) / length(!is.na(cbsa$CBSAFP))` percent of Black, and `r 100 * sum(cbsa$H_tenure_nhw > cbsa$H_inc_nhw_p50, na.rm = T) / length(!is.na(cbsa$CBSAFP))` percent of White observations. While there are quite a few metros where there is greater income ($H^R$) than tenure segregation within racial groups, especially Blacks, my hypothesis holds in a clear majority of metro areas and this result is even more clear with the like-for-like ($H(0.5)$) comparison. The following table confirms this result with a significance test similar to above.

```{r seg_diff_tests_race}
modelsummary(
  list("Tenure − Income H^R (NHW)"    = feols(I(H_tenure_nhw - H_inc_nhw)     ~ 1, cluster = ~CBSAFP + year, data = cbsa),
       "Tenure − Income H(0.5) (NHW)" = feols(I(H_tenure_nhw - H_inc_nhw_p50) ~ 1, cluster = ~CBSAFP + year, data = cbsa),
       "Tenure − Income H^R (NHB)"    = feols(I(H_tenure_nhb - H_inc_nhb)     ~ 1, cluster = ~CBSAFP + year, data = cbsa),
       "Tenure − Income H(0.5) (NHB)" = feols(I(H_tenure_nhb - H_inc_nhb_p50) ~ 1, cluster = ~CBSAFP + year, data = cbsa),
       "Tenure − Income H^R (His)"    = feols(I(H_tenure_his - H_inc_his)     ~ 1, cluster = ~CBSAFP + year, data = cbsa),
       "Tenure − Income H(0.5) (His)" = feols(I(H_tenure_his - H_inc_his_p50) ~ 1, cluster = ~CBSAFP + year, data = cbsa)),
  estimate    = "{estimate}{stars}",
  statistic   = c("({std.error})", "p {p.value}"),
  coef_rename = c("(Intercept)" = "Difference"),
  stars       = c("+" = 0.1, "*" = .05, "**" = .01, "***" = .001),
  gof_map     = c("nobs"),
  output      = "html"
)
```

The following table presents exact values of all segregation indices averaged over the 20 year period for each metro area for the interested reader.

```{r seg_table}
cbsa |>
  group_by(CBSAFP, CBSA_name) |>
  summarise(across(c(H_inc, H_inc_p50, H_rent, H_val, H_tenure,
                     H_inc_nhw, H_inc_nhw_p50, H_inc_nhb, H_inc_nhb_p50, H_inc_his, H_inc_his_p50,
                     H_tenure_nhw, H_tenure_nhb, H_tenure_his),
                   ~ mean(.x, na.rm = TRUE)),
            .groups = "drop") |>
  arrange(H_inc) |>
  select(CBSA_name, H_inc, H_inc_p50, H_rent, H_val, H_tenure,
         H_inc_nhw, H_inc_nhw_p50, H_tenure_nhw,
         H_inc_nhb, H_inc_nhb_p50, H_tenure_nhb,
         H_inc_his, H_inc_his_p50, H_tenure_his) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |>
  rename("Metro Area"       = CBSA_name,
         "Income H^R"       = H_inc,
         "Income H(0.5)"    = H_inc_p50,
         "Rent"             = H_rent,
         "Home Value"       = H_val,
         "Tenure"           = H_tenure,
         "Inc H^R (W)"      = H_inc_nhw,
         "Inc H(0.5) (W)"   = H_inc_nhw_p50,
         "Tenure (W)"       = H_tenure_nhw,
         "Inc H^R (B)"      = H_inc_nhb,
         "Inc H(0.5) (B)"   = H_inc_nhb_p50,
         "Tenure (B)"       = H_tenure_nhb,
         "Inc H^R (H)"      = H_inc_his,
         "Inc H(0.5) (H)"   = H_inc_his_p50,
         "Tenure (H)"       = H_tenure_his) |>
  knitr::kable(caption = "Segregation indices averaged across four ACS waves (2009–2024), ordered by overall income segregation") |>
  kable_styling(full_width = FALSE, font_size = 9, bootstrap_options = c("striped", "condensed")) |>
  add_header_above(c(" " = 1, "Overall" = 5, "NH-White" = 3, "NH-Black" = 3, "Hispanic" = 3))
```

# Segregation Moderates Income Inequality → Housing Inequality

The second city-level hypothesis I test here states that more unequal cities also show higher inequality in housing quality (proxied by rent / value). While it makes sense that a higher income inequality leads to a more unequal supply of housing quality because different people can afford different characteristics of housing. But my model states that investment decisions by landlords are contingent on local developments because landlords are side-ways looking under investment uncertainty. For the higher income inequality to materialize into higher inequality in housing, neighborhoods need to have different average incomes. Therefore, segregation moderates the relationship between income and rent inequality.

## Operationalisation of Housing Cost Inequality

The dependent variable is a Gini coefficient of housing costs pooled across renters and owners within each CBSA × year cell. Combining both tenure groups is necessary to capture two distinct dimensions of housing cost inequality simultaneously: *within-tenure* inequality (the spread of rents among renters, or of values among owners) and the *between-group* gap (the cost difference between renting and owning). A within-tenure-only measure — for instance, a renter-share-weighted average of separate Gini coefficients for rent and home value — is blind to the latter. Because the ABM predicts that segregation translates income inequality into housing inequality by sorting households into differentially desirable neighbourhoods regardless of tenure, the DV should reflect the full cross-tenure distribution of housing costs.

The central measurement challenge is that renters and owners face costs on incompatible scales: monthly gross rent for renters and total market value for owners. We convert home values to an imputed monthly user cost,

$$u_i = v_i \times \hat{r}_{mt} \;/\; 12,$$

where $v_i$ is the inflation-adjusted home value (`valueh_real`, 1999 USD) and $\hat{r}_{mt}$ is a metro- and wave-specific capitalisation rate derived from PUMS weighted medians:

$$\hat{r}_{mt} = 12 \;\times\; \widetilde{\mathrm{rent}}_{mt} \;/\; \widetilde{\mathrm{value}}_{mt}.$$

The factor of 12 follows from the standard real-estate definition of the annual gross yield: $r = \text{annual rent} / \text{value} = 12 \times \text{monthly rent} / \text{value}$. Dividing by 12 in the user-cost formula then recovers a monthly figure, so the two factors cancel and the imputed cost simplifies to $u_i = v_i \times \widetilde{\mathrm{rent}}_{mt} / \widetilde{\mathrm{value}}_{mt}$. Both medians are computed from the same PUMS records used for the Gini, using household survey weights (HHWT). Cross-metro variation in $\hat{r}$ is substantively meaningful — the rent-to-price ratio is systematically lower in supply-constrained coastal metros than in Rust Belt markets — and should be preserved rather than suppressed by a fixed national rate.

**Key assumption.** The formula embeds the assumption that rent is *proportional* to value across the full within-metro quality distribution: a home worth twice the median is assumed to rent for twice the median rent. Equivalently, the cap rate is treated as constant across quality tiers within a metro. This is a standard first-order approximation but is unlikely to hold exactly: in most markets, luxury units command a lower gross yield than entry-level units (i.e., the cap rate declines with quality), which means the formula overstates imputed costs at the top of the value distribution and understates them at the bottom. The resulting Gini will therefore be somewhat compressed relative to the true housing cost distribution. This bias cannot be eliminated without observing what owner-occupied units would rent for in the current market, which is not available in ACS or PUMS. The pooled monthly housing cost is then

$$c_i = \begin{cases} \text{rentgrs\_real}_i & \text{renter} \\ v_i \times \hat{r}_{mt} / 12 & \text{owner,} \end{cases}$$

and `gini_housing_cbsa` is the HHWT-weighted Gini of $c_i$ within each CBSA × year (Brown formula, `wtd_gini` in `SEHTestACS3_citylevel.R`).

**Robustness checks on the capitalisation rate.** The metro-specific $\hat{r}_{mt}$ could confound the Gini if the rent-to-price ratio itself co-varies with income inequality or segregation, creating a spurious correlation rather than a genuine measurement of housing cost inequality. Three alternative cap rate assumptions are therefore carried as robustness checks:

| Variant | Cap rate | What it isolates |
|---|---|---|
| `gini_housing_cbsa` | Metro×year from PUMS medians | Main specification |
| `gini_housing_cbsa_5pct` | Fixed 5 % p.a. | Standard housing-economics benchmark |
| `gini_housing_cbsa_7pct` | Fixed 7 % p.a. | Higher cost-of-capital assumption |
| `gini_housing_cbsa_natl` | Year-specific national rate | Removes cross-metro variation in cap rates |

The year-specific national rate is particularly informative: if estimates are stable when switching from `gini_housing_cbsa` to `gini_housing_cbsa_natl`, cross-metro differences in rent-to-price ratios are not driving the results.

## Test of the Hypothesis

```{r housing_ineq_models}
# Mean-centre continuous predictors so main effects are evaluated at the sample
# mean of the moderator rather than at zero (which is outside the observed range)
cbsa <- cbsa |>
  mutate(
    gini_inc_c = gini_inc_cbsa - mean(gini_inc_cbsa, na.rm = TRUE),
    H_inc_c    = H_inc         - mean(H_inc,         na.rm = TRUE)
  )

cm1 <- feols(gini_housing_cbsa ~ gini_inc_c | year,
             data = cbsa, cluster = ~CBSAFP + year)
cm2 <- feols(gini_housing_cbsa ~ gini_inc_c * H_inc_c | year,
             data = cbsa, cluster = ~CBSAFP + year)
cm3 <- feols(gini_housing_cbsa ~ gini_inc_c * H_inc_c + H_bw +
               unemp_rate_cbsa + vacancy_rate_cbsa + pct_renter_cbsa | year,
             data = cbsa, cluster = ~CBSAFP + year)

coef_names <- c(
  "gini_inc_c"         = "Income Gini (centred)",
  "H_inc_c"            = "Income Segregation H^R (centred)",
  "gini_inc_c:H_inc_c" = "Income Gini × H^R",
  "H_bw"               = "Black-White Segregation H",
  "unemp_rate_cbsa"    = "Unemployment Rate",
  "vacancy_rate_cbsa"  = "Vacancy Rate",
  "pct_renter_cbsa"    = "Renter Share"
)

modelsummary(
  list("(1)" = cm1, "(2)" = cm2, "(3)" = cm3),
  estimate    = "{estimate}{stars}",
  statistic   = "({std.error})",
  coef_rename = coef_names,
  stars       = c("+" = 0.1, "*" = .05, "**" = .01, "***" = .001),
  gof_map     = c("nobs", "r.squared", "FE: year"),
  output      = "html"
)
```

The results consistently show a positive and significant main effect of income inequality on housing cost inequality. The moderation hypothesis — that income segregation amplifies this relationship — receives directional support (positive interaction) when adding the control variables but does not reach significance.

The year-FE specification is the theoretically appropriate primary test: the ABM generates cross-sectional predictions by comparing equilibria across cities that differ in their level of segregation. In principle, within-metro changes in segregation would also constitute a valid test of the mechanism — if neighbourhoods diverge in their income composition, the ABM predicts that developers and landlords will respond by investing differently across places, widening the housing cost distribution. However, two features of the setting work against detecting this effect. First, residential segregation is highly stable within cities over medium-run horizons — changes in H^R within any single metro across the four waves are likely small relative to cross-metro differences. Second, even genuine increases in segregation would translate into housing cost inequality only with a substantial lag: upgrading of desirable neighbourhoods, decay of neglected ones, and new construction all operate on timescales of years to decades that the 5-year inter-wave gaps cannot capture. The first-differences model below addresses this directly.

## Robustness: Cap Rate Sensitivity

Model (3) re-estimated with the three alternative cap rates. Stability of the interaction coefficient across specifications rules out that the rent-to-price ratio assumption is driving the moderation result.

```{r housing_ineq_robustness}
modelsummary(
  list(
    "Metro"    = cm3,
    "Fixed 5%" = feols(gini_housing_cbsa_5pct ~ gini_inc_c * H_inc_c + H_bw +
                         unemp_rate_cbsa + vacancy_rate_cbsa + pct_renter_cbsa | year,
                       data = cbsa, cluster = ~CBSAFP + year),
    "Fixed 7%" = feols(gini_housing_cbsa_7pct ~ gini_inc_c * H_inc_c + H_bw +
                         unemp_rate_cbsa + vacancy_rate_cbsa + pct_renter_cbsa | year,
                       data = cbsa, cluster = ~CBSAFP + year),
    "National" = feols(gini_housing_cbsa_natl ~ gini_inc_c * H_inc_c + H_bw +
                         unemp_rate_cbsa + vacancy_rate_cbsa + pct_renter_cbsa | year,
                       data = cbsa, cluster = ~CBSAFP + year)
  ),
  estimate    = "{estimate}{stars}",
  statistic   = "({std.error})",
  coef_rename = coef_names,
  #coef_omit   = "^(?!.*gini_inc_c|.*H_inc_c)",
  stars       = c("+" = 0.1, "*" = .05, "**" = .01, "***" = .001),
  gof_map     = c("nobs", "r.squared", "FE: year"),
  output      = "html"
)
```

## First-Differences Model

The levels model tests whether more segregated cities have higher housing cost inequality. The first-differences (FD) model asks a different question: does a rise in income inequality translate into a larger rise in housing costs in more segregated cities? This directly tests the temporal mechanism — segregation shaping how income inequality propagates through developer and landlord investment decisions over time — without the between-metro confounding that motivates the concern about the year-FE estimates.

First-differencing removes all time-constant between-city heterogeneity (including persistently high-demand metros vs. stagnating ones) without requiring unit FE and therefore without Nickell bias. Three design choices follow from the nature of the estimand. First, the DV and main predictor are differenced ($\Delta$), but income segregation enters as its *lagged level* $H^R_{m,t-5}$: it is the structural context within which income inequality changes play out, it is predetermined relative to $\Delta\text{GiniHousing}$, and as established above it barely moves within metros over the sample period. Second, all time-varying controls enter as period changes ($\Delta$) to match the estimand and absorb demand and economic activity shocks that could confound the interaction — in particular changes in population growth, unemployment, vacancy, renter share, and racial segregation. Population growth (`pop_growth_cbsa`) is already computed as a period growth rate in `SEHTestACS3_citylevel.R` and enters directly. Third, year FE are retained as two dummies for the three inter-wave difference periods (2009→2014, 2014→2019, 2019→2024) to absorb national trends in housing cost growth. Standard errors are clustered by CBSAFP only rather than two-way, because the FD sample contains only three period clusters — well below the threshold for reliable two-way inference — and the year FE already absorb systematic inter-period shifts, leaving within-CBSA serial correlation as the dominant source of dependence. Segregation $H^R_{m,t-5}$ is mean-centred so that the main effect of $\Delta\text{GiniInc}$ is evaluated at the average metro.

```{r housing_ineq_fd}
cbsa_fd <- cbsa |>
  arrange(CBSAFP, year) |>
  group_by(CBSAFP) |>
  mutate(
    d_gini_housing      = gini_housing_cbsa      - lag(gini_housing_cbsa),
    d_gini_housing_5pct = gini_housing_cbsa_5pct - lag(gini_housing_cbsa_5pct),
    d_gini_housing_7pct = gini_housing_cbsa_7pct - lag(gini_housing_cbsa_7pct),
    d_gini_housing_natl = gini_housing_cbsa_natl - lag(gini_housing_cbsa_natl),
    d_gini_rent         = gini_rent_cbsa         - lag(gini_rent_cbsa),
    d_gini_val          = gini_val_cbsa          - lag(gini_val_cbsa),
    d_gini_inc          = gini_inc_cbsa          - lag(gini_inc_cbsa),
    d_H_bw              = H_bw                   - lag(H_bw),
    d_unemp             = unemp_rate_cbsa        - lag(unemp_rate_cbsa),
    d_vacancy           = vacancy_rate_cbsa      - lag(vacancy_rate_cbsa),
    d_pct_renter        = pct_renter_cbsa        - lag(pct_renter_cbsa),
    H_inc_lag           = lag(H_inc)
  ) |>
  ungroup() |>
  filter(!is.na(d_gini_housing)) |>
  mutate(H_inc_lag_c = H_inc_lag - mean(H_inc_lag, na.rm = TRUE))

fd_coef_names <- c(
  "d_gini_inc"             = "\u0394Income Gini",
  "H_inc_lag_c"            = "Income Segregation H\u1d3f (t\u22125, centred)",
  "d_gini_inc:H_inc_lag_c" = "\u0394Income Gini \u00d7 H\u1d3f (t\u22125)",
  "d_H_bw"                 = "\u0394Black-White Segregation H",
  "d_unemp"                = "\u0394Unemployment Rate",
  "d_vacancy"              = "\u0394Vacancy Rate",
  "d_pct_renter"           = "\u0394Renter Share",
  "pop_growth_cbsa"        = "Population Growth"
)

cm_fd1 <- feols(d_gini_housing ~ d_gini_inc | year,
                data = cbsa_fd, cluster = ~CBSAFP)
cm_fd2 <- feols(d_gini_housing ~ d_gini_inc * H_inc_lag_c | year,
                data = cbsa_fd, cluster = ~CBSAFP)
cm_fd3 <- feols(d_gini_housing ~ d_gini_inc * H_inc_lag_c + d_H_bw +
                  d_unemp + d_vacancy + d_pct_renter + pop_growth_cbsa | year,
                data = cbsa_fd, cluster = ~CBSAFP)

modelsummary(
  list("FD 1" = cm_fd1, "FD 2" = cm_fd2, "FD 3" = cm_fd3),
  estimate    = "{estimate}{stars}",
  statistic   = "({std.error})",
  coef_rename = fd_coef_names,
  stars       = c("+" = 0.1, "*" = .05, "**" = .01, "***" = .001),
  gof_map     = c("nobs", "r.squared", "FE: year"),
  output      = "html"
)
```

The main effect of income inequality changes ($\Delta$Income Gini) on housing cost inequality changes is positive and consistent across models, mirroring the levels result. The interaction between $\Delta$Income Gini and lagged segregation $H^R_{t-5}$ is positive and substantively large, consistent with the theoretical prediction that more segregated cities translate rising income inequality into rising housing cost inequality more strongly. It does not, however, reach conventional significance thresholds.

Two features of the FD design limit statistical power here. First, first-differencing removes the between-metro variance that constitutes the bulk of variation in both income inequality and segregation, leaving only the modest within-metro, period-to-period changes to identify the interaction. Second, with only three difference periods per metro and one-way CBSAFP clustering, degrees of freedom for the year dimension are exhausted, making the standard errors conservative. The positive direction of the interaction is nonetheless consistent across the trivariate model and the full specification with change controls — including controls specifically designed to absorb the alternative explanation that high-segregation metros are simply faster-growing cities with rising housing demand. The effect is slightly reduced in the model with controls but is still sizeable. The consistency of the sign across both the levels and FD models constitutes cautious support for the moderation hypothesis.

### Predicted Values

The plot below illustrates the moderation: predicted $\Delta$Housing Cost Gini as a function of $\Delta$Income Gini at three levels of lagged income segregation (low = $-1$ SD, mean = 0, high = $+1$ SD). All other predictors are held at zero (a no-change baseline). Year FE are absorbed. The steeper slope for high-segregation metros captures the predicted amplification effect.

```{r housing_ineq_fd_pred, fig.height=4}
b      <- coef(cm_fd3)
V      <- vcov(cm_fd3)[c("d_gini_inc", "H_inc_lag_c", "d_gini_inc:H_inc_lag_c"),
                        c("d_gini_inc", "H_inc_lag_c", "d_gini_inc:H_inc_lag_c")]
h_sd   <- sd(cbsa_fd$H_inc_lag_c, na.rm = TRUE)
x_seq  <- seq(min(cbsa_fd$d_gini_inc, na.rm = TRUE),
              max(cbsa_fd$d_gini_inc, na.rm = TRUE),
              length.out = 100)

h_levels <- c(-h_sd, 0, h_sd)
h_labels <- c("Low ($-$1 SD)", "Mean", "High ($+$1 SD)")

expand.grid(d_gini_inc = x_seq, H_inc_lag_c = h_levels) |>
  mutate(
    H_label = factor(H_inc_lag_c, levels = h_levels, labels = h_labels),
    y_hat   = b["d_gini_inc"]             * d_gini_inc
            + b["H_inc_lag_c"]            * H_inc_lag_c
            + b["d_gini_inc:H_inc_lag_c"] * d_gini_inc * H_inc_lag_c,
    se      = mapply(function(x, h) {
                z <- c(x, h, x * h)
                sqrt(as.numeric(z %*% V %*% z))
              }, d_gini_inc, H_inc_lag_c),
    ylo     = y_hat - 1.96 * se,
    yhi     = y_hat + 1.96 * se
  ) |>
  ggplot(aes(x = d_gini_inc, colour = H_label, fill = H_label, linetype = H_label)) +
  geom_ribbon(aes(ymin = ylo, ymax = yhi), alpha = 0.15, colour = NA) +
  geom_line(aes(y = y_hat), linewidth = 0.8) +
  scale_colour_manual(values = cb_palette[c(1, 3, 2)],
                      name   = TeX("Income Segregation $H^R$"),
                      labels = lapply(h_labels, TeX)) +
  scale_fill_manual(values   = cb_palette[c(1, 3, 2)],
                    name     = TeX("Income Segregation $H^R$"),
                    labels   = lapply(h_labels, TeX)) +
  scale_linetype_manual(values = lines[1:3],
                        name   = TeX("Income Segregation $H^R$"),
                        labels = lapply(h_labels, TeX)) +
  labs(
    x       = TeX("$\\Delta$ Income Gini"),
    y       = TeX("$\\Delta$ Housing Cost Gini")
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
ggsave(paste0(results_dir, "seg_ineqFD.pdf"), height = 12, width = 18, units = "cm")
```

Income inequality also changes little over time within metros. The range in delta Gini over five years is only about 0.05.


### Cap Rate Robustness

FD 3 re-estimated with the three alternative cap rates, showing only the key coefficients.

```{r housing_ineq_fd_rob}
modelsummary(
  list(
    "Metro"    = cm_fd3,
    "Fixed 5%" = feols(d_gini_housing_5pct ~ d_gini_inc * H_inc_lag_c + d_H_bw +
                         d_unemp + d_vacancy + d_pct_renter + pop_growth_cbsa | year,
                       data = cbsa_fd, cluster = ~CBSAFP),
    "Fixed 7%" = feols(d_gini_housing_7pct ~ d_gini_inc * H_inc_lag_c + d_H_bw +
                         d_unemp + d_vacancy + d_pct_renter + pop_growth_cbsa | year,
                       data = cbsa_fd, cluster = ~CBSAFP),
    "National" = feols(d_gini_housing_natl ~ d_gini_inc * H_inc_lag_c + d_H_bw +
                         d_unemp + d_vacancy + d_pct_renter + pop_growth_cbsa | year,
                       data = cbsa_fd, cluster = ~CBSAFP)
  ),
  estimate    = "{estimate}{stars}",
  statistic   = "({std.error})",
  coef_rename = fd_coef_names,
  #coef_omit   = "^(?!.*d_gini_inc|.*H_inc_lag_c)",
  stars       = c("+" = 0.1, "*" = .05, "**" = .01, "***" = .001),
  gof_map     = c("nobs", "r.squared", "FE: year"),
  output      = "html"
)
```

## Decomposition by Tenure

The pooled Gini combines within-tenure inequality with the renter/owner cost gap. To understand which component drives the moderation result, we re-estimate the full model separately using `gini_rent_cbsa` (renters only) and `gini_val_cbsa` (owners only) as dependent variables. Because these measures are computed from directly observed values without any cap rate conversion, this also bypasses the imputed rent assumption entirely.

Recall that the moderation hypothesis concerns how strongly a rise in income inequality translates into housing cost inequality given the level of segregation — segregation is the moderator, not the driver. With this in mind, we expect the moderation to be stronger for home values than for rents. When income inequality increases (especially at the top), in more segregated metros the gains of top earners flow into a narrow set of premium neighborhoods where high-income households are concentrated, bidding up values there. At the same time, Glaeser and Gyourko's (2005) supply asymmetry implies that low-income areas do not depreciate symmetrically — the durable housing stock decays slowly, creating a floor under low-end values. Together, concentrated appreciation at the top and sticky values at the bottom amplify how income inequality growth translates into value dispersion, more so in metros where income groups are spatially separated. The rental market absorbs the same income inequality increase more diffusely: renters are more mobile across neighbourhoods, lease contracts slow price adjustment, and the income–rent gradient is less steep to begin with.

Note that null results here do not contradict the main finding: if the moderation operates primarily through the between-group channel — segregation widening the cost gap between renters and owners as groups — it would not appear in either tenure-specific Gini.

```{r housing_ineq_decomp}
modelsummary(
  list(
    "Rent"  = feols(gini_rent_cbsa ~ gini_inc_c * H_inc_c + H_bw +
                      unemp_rate_cbsa + vacancy_rate_cbsa | year,
                    data = cbsa, cluster = ~CBSAFP + year),
    "Value" = feols(gini_val_cbsa  ~ gini_inc_c * H_inc_c + H_bw +
                      unemp_rate_cbsa + vacancy_rate_cbsa | year,
                    data = cbsa, cluster = ~CBSAFP + year)
  ),
  estimate    = "{estimate}{stars}",
  statistic   = "({std.error})",
  coef_rename = coef_names,
  coef_omit   = "^(?!.*gini_inc_c|.*H_inc_c)",
  stars       = c("+" = 0.1, "*" = .05, "**" = .01, "***" = .001),
  gof_map     = c("nobs", "r.squared", "FE: year"),
  output      = "html"
)
```

### First-Differences Decomposition

```{r housing_ineq_fd_decomp}
modelsummary(
  list(
    "\u0394Rent (FD)"  = feols(d_gini_rent ~ d_gini_inc * H_inc_lag_c + d_H_bw +
                                 d_unemp + d_vacancy | year,
                               data = cbsa_fd, cluster = ~CBSAFP),
    "\u0394Value (FD)" = feols(d_gini_val  ~ d_gini_inc * H_inc_lag_c + d_H_bw +
                                 d_unemp + d_vacancy | year,
                               data = cbsa_fd, cluster = ~CBSAFP)
  ),
  estimate    = "{estimate}{stars}",
  statistic   = "({std.error})",
  coef_rename = fd_coef_names,
  stars       = c("+" = 0.1, "*" = .05, "**" = .01, "***" = .001),
  gof_map     = c("nobs", "r.squared", "FE: year"),
  output      = "html"
)
```

The FD results confirm the asymmetry: the interaction effect for home values is substantially larger than for rents and reaches marginal significance, while the rent interaction is small and insignificant. This is consistent with the theoretical expectation above. The FD estimator identifies changes within metros over time, so the value result specifically says that in metros where income segregation was higher at the start of a period, a given increase in income inequality over that period translated into a larger widening of home value dispersion. This dynamic pattern fits the capitalisation story: rising top incomes bid up values in the premium (segregated) neighbourhoods disproportionately, while value floors in low-income areas are held up by durable housing stock. Rents, by contrast, adjust more slowly and across a more diffuse set of neighbourhoods, dampening the within-metro, within-period translation.


# Segregation Moderates Income → Housing Quality

The ABM predicts that income segregation acts as a moderator of the relationship between household income and housing quality. In more segregated cities, neighbourhoods are more polarised in their attractiveness and housing quality. High-income households cluster in high-quality areas while low-income households are confined to low-quality ones, steepening the income–quality gradient. The cross-level interaction hypothesis is:

> In more segregated cities, the effect of household income on housing quality is stronger (steeper slope).

We test this with a WLS regression of within-group quality rank on log equivalised household income, the city-level rank-order income segregation index $H^R$, and their interaction. Household weights (HHWT) are applied throughout. Standard errors are clustered by CBSA. Two-way clustering on CBSA and year is not used: with only four survey waves the year dimension has far too few clusters for reliable cluster-robust inference.

No fixed effects are included: `quality` is already z-standardised within `YEAR × CBSAFP × tenure`, which removes all cell means — a finer demeaning than any combination of additive year and CBSA FEs. The identifying variation is therefore already purely within-cell (between households in the same metro, year, and tenure group). Instead, we control directly for the following time-varying individual and metro-level confounders:

- metro area income inequality (Gini index): 
- metro area racial segregation (Theil index White vs. Black):
- metro area unemployment rate:
- metro area vacancy rate: 
- individual race/ethnicity:
- individual education (college):
- individual age and age squared: 


```{r quality_combined}
m1 <- feols(quality ~ log_equiv_inc + race_eth + college + AGE + I(AGE^2) +
              gini_inc_cbsa + H_bw + unemp_rate_cbsa + vacancy_rate_cbsa,
            data = pums, weights = ~HHWT, cluster = "CBSAFP")
m2 <- feols(quality ~ log_equiv_inc + H_inc + race_eth + college + AGE + I(AGE^2) +
              gini_inc_cbsa + H_bw + unemp_rate_cbsa + vacancy_rate_cbsa,
            data = pums, weights = ~HHWT, cluster = "CBSAFP")
m3 <- feols(quality ~ log_equiv_inc * H_inc + race_eth + college + AGE + I(AGE^2) +
              gini_inc_cbsa + H_bw + unemp_rate_cbsa + vacancy_rate_cbsa,
            data = pums, weights = ~HHWT, cluster = "CBSAFP")

modelsummary(
  list("(1)" = m1, "(2)" = m2, "(3)" = m3),
  stars   = c("+" = 0.1, "*" = .05, "**" = .01, "***" = .001),
  gof_map = c("nobs", "r.squared"),
  output  = "html"
)
```

```{r quality_combined_plot}
b    <- coef(m3)
V    <- vcov(m3)

age_mn   <- weighted.mean(pums$AGE,               pums$HHWT, na.rm = TRUE)
gini_mn  <- weighted.mean(pums$gini_inc_cbsa,     pums$HHWT, na.rm = TRUE)
Hbw_mn   <- weighted.mean(pums$H_bw,              pums$HHWT, na.rm = TRUE)
unemp_mn <- weighted.mean(pums$unemp_rate_cbsa,   pums$HHWT, na.rm = TRUE)
vac_mn   <- weighted.mean(pums$vacancy_rate_cbsa, pums$HHWT, na.rm = TRUE)
h_mn     <- weighted.mean(pums$H_inc,             pums$HHWT, na.rm = TRUE)
h_sd     <- sd(pums$H_inc, na.rm = TRUE)

x_seq    <- seq(quantile(pums$log_equiv_inc, 0.01, na.rm = TRUE),
                quantile(pums$log_equiv_inc, 0.99, na.rm = TRUE),
                length.out = 100)
h_levels <- h_mn + c(-h_sd, 0, h_sd)
h_labels <- c("Low ($-$1 SD)", "Mean", "High ($+$1 SD)")

newdata <- expand.grid(
  log_equiv_inc     = x_seq,
  H_inc             = h_levels,
  race_eth          = factor("nhwhite", levels = levels(pums$race_eth)),
  college           = 0,
  AGE               = age_mn,
  gini_inc_cbsa     = gini_mn,
  H_bw              = Hbw_mn,
  unemp_rate_cbsa   = unemp_mn,
  vacancy_rate_cbsa = vac_mn
) |>
  mutate(H_label = factor(H_inc, levels = h_levels, labels = h_labels))

V_key <- V[c("log_equiv_inc", "H_inc", "log_equiv_inc:H_inc"),
            c("log_equiv_inc", "H_inc", "log_equiv_inc:H_inc")]

newdata <- newdata |>
  mutate(
    y_hat = predict(m3, newdata = newdata),
    se    = mapply(function(x, h) {
              z <- c(x, h, x * h)
              sqrt(as.numeric(z %*% V_key %*% z))
            }, log_equiv_inc, H_inc),
    ylo   = y_hat - 1.96 * se,
    yhi   = y_hat + 1.96 * se
  )

ggplot(newdata, aes(x = exp(log_equiv_inc),
                    colour = H_label, fill = H_label, linetype = H_label)) +
  geom_ribbon(aes(ymin = ylo, ymax = yhi), alpha = 0.15, colour = NA) +
  geom_line(aes(y = y_hat), linewidth = 0.8) +
  scale_colour_manual(values = cb_palette[c(1, 3, 2)],
                      name   = TeX("Income Segregation $H^R$"),
                      labels = lapply(h_labels, TeX)) +
  scale_fill_manual(values   = cb_palette[c(1, 3, 2)],
                    name     = TeX("Income Segregation $H^R$"),
                    labels   = lapply(h_labels, TeX)) +
  scale_linetype_manual(values = lines[1:3],
                        name   = TeX("Income Segregation $H^R$"),
                        labels = lapply(h_labels, TeX)) +
  scale_x_continuous(labels = scales::label_dollar()) +
  labs(
    x = "Equivalised Household Income (1999 USD)",
    y = "Housing Quality (predicted z-score)"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
ggsave(paste0(results_dir, "quality_income_seg.pdf"), height = 12, width = 18, units = "cm")
```


## By Tenure

The income–quality relationship may operate through different mechanisms for renters and owners. For renters, income determines which segment of the rental market a household can access. For owners, income determines purchasing power in the sales market, where neighbourhood sorting is reinforced by credit constraints and property tax capitalisation of local amenities. We therefore estimate the interaction model separately by tenure.

```{r quality_tenure}
pums_renters <- filter(pums, renter == 1L)
pums_owners  <- filter(pums, renter == 0L)

m_r1 <- feols(quality ~ log_equiv_inc + race_eth + college + AGE + I(AGE^2) +
                gini_inc_cbsa + H_bw + unemp_rate_cbsa + vacancy_rate_cbsa,
              data = pums_renters, weights = ~HHWT, cluster = "CBSAFP")
m_r2 <- feols(quality ~ log_equiv_inc + H_inc + race_eth + college + AGE + I(AGE^2) +
                gini_inc_cbsa + H_bw + unemp_rate_cbsa + vacancy_rate_cbsa,
              data = pums_renters, weights = ~HHWT, cluster = "CBSAFP")
m_r3 <- feols(quality ~ log_equiv_inc * H_inc + race_eth + college + AGE + I(AGE^2) +
                gini_inc_cbsa + H_bw + unemp_rate_cbsa + vacancy_rate_cbsa,
              data = pums_renters, weights = ~HHWT, cluster = "CBSAFP")

m_o1 <- feols(quality ~ log_equiv_inc + race_eth + college + AGE + I(AGE^2) +
                gini_inc_cbsa + H_bw + unemp_rate_cbsa + vacancy_rate_cbsa,
              data = pums_owners,  weights = ~HHWT, cluster = "CBSAFP")
m_o2 <- feols(quality ~ log_equiv_inc + H_inc + race_eth + college + AGE + I(AGE^2) +
                gini_inc_cbsa + H_bw + unemp_rate_cbsa + vacancy_rate_cbsa,
              data = pums_owners,  weights = ~HHWT, cluster = "CBSAFP")
m_o3 <- feols(quality ~ log_equiv_inc * H_inc + race_eth + college + AGE + I(AGE^2) +
                gini_inc_cbsa + H_bw + unemp_rate_cbsa + vacancy_rate_cbsa,
              data = pums_owners,  weights = ~HHWT, cluster = "CBSAFP")

modelsummary(
  list("Renters (1)" = m_r1, "Renters (2)" = m_r2, "Renters (3)" = m_r3,
       "Owners (1)"  = m_o1, "Owners (2)"  = m_o2, "Owners (3)"  = m_o3),
  stars   = c("+" = 0.1, "*" = .05, "**" = .01, "***" = .001),
  gof_map = c("nobs", "r.squared"),
  output  = "html"
)
```

# Segregation Moderates Income → Housing Cost Burden

> In more segregated cities, the effect of household income on housing cost burden is weaker (softer slope).

Because the outcome variables are not pre-standardised, year and CBSAFP fixed effects are included throughout. Although `H_inc` changes little within metros over the sample period, it does vary, and the cross-level interaction `log_equiv_inc × H_inc` remains identified with CBSAFP FE: the interaction coefficient captures how the within-household income slope on cost burden varies with metro-level segregation, using both cross-sectional and within-metro over-time variation. The main effect of `H_inc` is absorbed by the CBSAFP FE and not interpretable in models that include it. Standard errors are clustered by CBSAFP; two-way clustering on year is not used because only four survey waves are available.

The control strategy mirrors the quality models: individual controls (race/ethnicity, college degree, age and age squared) and time-varying metro-level controls (income inequality Gini, Black-White segregation H, unemployment rate, vacancy rate).

## Renters

The renter outcome is the annual rent-to-income ratio `rent_to_inc` = (RENTGRS × 12) / HHINCOME. RENTGRS is monthly gross rent; multiplying by 12 annualises it to match HHINCOME (annual). The CPI deflator cancels.

```{r burden_models}
pums_renters_b <- filter(pums, !is.na(rent_to_inc))

b1 <- feols(rent_to_inc ~ log_equiv_inc + race_eth + college + AGE + I(AGE^2) +
              gini_inc_cbsa + H_bw + unemp_rate_cbsa + vacancy_rate_cbsa | YEAR + CBSAFP,
            data = pums_renters_b, weights = ~HHWT, cluster = "CBSAFP")
b2 <- feols(rent_to_inc ~ log_equiv_inc + H_inc + race_eth + college + AGE + I(AGE^2) +
              gini_inc_cbsa + H_bw + unemp_rate_cbsa + vacancy_rate_cbsa | YEAR + CBSAFP,
            data = pums_renters_b, weights = ~HHWT, cluster = "CBSAFP")
b3 <- feols(rent_to_inc ~ log_equiv_inc * H_inc + race_eth + college + AGE + I(AGE^2) +
              gini_inc_cbsa + H_bw + unemp_rate_cbsa + vacancy_rate_cbsa | YEAR + CBSAFP,
            data = pums_renters_b, weights = ~HHWT, cluster = "CBSAFP")

modelsummary(
  list("(1)" = b1, "(2)" = b2, "(3)" = b3),
  estimate    = "{estimate}{stars}",
  statistic   = "({std.error})",
  stars       = c("+" = 0.1, "*" = .05, "**" = .01, "***" = .001),
  gof_map     = c("nobs", "r.squared", "FE: YEAR", "FE: CBSAFP"),
  output      = "html"
)
```

```{r burden_plot}
b_coef <- coef(b3)
b_vcov <- vcov(b3)

age_mn   <- weighted.mean(pums_renters_b$AGE,               pums_renters_b$HHWT, na.rm = TRUE)
gini_mn  <- weighted.mean(pums_renters_b$gini_inc_cbsa,     pums_renters_b$HHWT, na.rm = TRUE)
Hbw_mn   <- weighted.mean(pums_renters_b$H_bw,              pums_renters_b$HHWT, na.rm = TRUE)
unemp_mn <- weighted.mean(pums_renters_b$unemp_rate_cbsa,   pums_renters_b$HHWT, na.rm = TRUE)
vac_mn   <- weighted.mean(pums_renters_b$vacancy_rate_cbsa, pums_renters_b$HHWT, na.rm = TRUE)
h_mn     <- weighted.mean(pums_renters_b$H_inc,             pums_renters_b$HHWT, na.rm = TRUE)
h_sd     <- sd(pums_renters_b$H_inc, na.rm = TRUE)

x_seq    <- seq(quantile(pums_renters_b$log_equiv_inc, 0.01, na.rm = TRUE),
                quantile(pums_renters_b$log_equiv_inc, 0.99, na.rm = TRUE),
                length.out = 100)
h_levels <- h_mn + c(-h_sd, 0, h_sd)
h_labels <- c("Low ($-$1 SD)", "Mean", "High ($+$1 SD)")

newdata_b <- expand.grid(
  log_equiv_inc     = x_seq,
  H_inc             = h_levels,
  race_eth          = factor("nhwhite", levels = levels(pums$race_eth)),
  college           = 0,
  AGE               = age_mn,
  gini_inc_cbsa     = gini_mn,
  H_bw              = Hbw_mn,
  unemp_rate_cbsa   = unemp_mn,
  vacancy_rate_cbsa = vac_mn
) |>
  mutate(H_label = factor(H_inc, levels = h_levels, labels = h_labels))

V_b_key <- b_vcov[c("log_equiv_inc", "H_inc", "log_equiv_inc:H_inc"),
                   c("log_equiv_inc", "H_inc", "log_equiv_inc:H_inc")]

newdata_b <- newdata_b |>
  mutate(
    y_hat = predict(b3, newdata = newdata_b),
    se    = mapply(function(x, h) {
              z <- c(x, h, x * h)
              sqrt(as.numeric(z %*% V_b_key %*% z))
            }, log_equiv_inc, H_inc),
    ylo   = y_hat - 1.96 * se,
    yhi   = y_hat + 1.96 * se
  )

ggplot(newdata_b, aes(x = exp(log_equiv_inc),
                      colour = H_label, fill = H_label, linetype = H_label)) +
  geom_ribbon(aes(ymin = ylo, ymax = yhi), alpha = 0.15, colour = NA) +
  geom_line(aes(y = y_hat), linewidth = 0.8) +
  scale_colour_manual(values = cb_palette[c(1, 3, 2)],
                      name   = TeX("Income Segregation $H^R$"),
                      labels = lapply(h_labels, TeX)) +
  scale_fill_manual(values   = cb_palette[c(1, 3, 2)],
                    name     = TeX("Income Segregation $H^R$"),
                    labels   = lapply(h_labels, TeX)) +
  scale_linetype_manual(values = lines[1:3],
                        name   = TeX("Income Segregation $H^R$"),
                        labels = lapply(h_labels, TeX)) +
  scale_x_continuous(labels = scales::label_dollar()) +
  labs(
    x = "Equivalised Household Income (1999 USD)",
    y = "Rent-to-Income Ratio (predicted)"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
ggsave(paste0(results_dir, "burden_income_seg.pdf"), height = 12, width = 18, units = "cm")
```

## Owners

For owners two cost proxies are available. The value-to-income ratio (`val_to_inc` = `valueh_real / hhincome_real`) is a stock/flow measure directly available from the PUMS. The imputed cost-to-income ratio (`cost_to_inc` = `valueh_real × cap_rate_metro / hhincome_real`) converts the property value to an annualised flow cost using the metro × year cap rate, giving a flow/flow ratio directly comparable to the renter measure. Both are deflator-invariant.

### Value-to-Income Ratio

```{r burden_val_models}
pums_owners_v <- filter(pums, !is.na(val_to_inc))

v1 <- feols(val_to_inc ~ log_equiv_inc + race_eth + college + AGE + I(AGE^2) +
              gini_inc_cbsa + H_bw + unemp_rate_cbsa + vacancy_rate_cbsa | YEAR + CBSAFP,
            data = pums_owners_v, weights = ~HHWT, cluster = "CBSAFP")
v2 <- feols(val_to_inc ~ log_equiv_inc + H_inc + race_eth + college + AGE + I(AGE^2) +
              gini_inc_cbsa + H_bw + unemp_rate_cbsa + vacancy_rate_cbsa | YEAR + CBSAFP,
            data = pums_owners_v, weights = ~HHWT, cluster = "CBSAFP")
v3 <- feols(val_to_inc ~ log_equiv_inc * H_inc + race_eth + college + AGE + I(AGE^2) +
              gini_inc_cbsa + H_bw + unemp_rate_cbsa + vacancy_rate_cbsa | YEAR + CBSAFP,
            data = pums_owners_v, weights = ~HHWT, cluster = "CBSAFP")

modelsummary(
  list("(1)" = v1, "(2)" = v2, "(3)" = v3),
  estimate    = "{estimate}{stars}",
  statistic   = "({std.error})",
  stars       = c("+" = 0.1, "*" = .05, "**" = .01, "***" = .001),
  gof_map     = c("nobs", "r.squared", "FE: YEAR", "FE: CBSAFP"),
  output      = "html"
)
```

```{r burden_val_plot}
v_coef <- coef(v3)
v_vcov <- vcov(v3)

age_mn   <- weighted.mean(pums_owners_v$AGE,               pums_owners_v$HHWT, na.rm = TRUE)
gini_mn  <- weighted.mean(pums_owners_v$gini_inc_cbsa,     pums_owners_v$HHWT, na.rm = TRUE)
Hbw_mn   <- weighted.mean(pums_owners_v$H_bw,              pums_owners_v$HHWT, na.rm = TRUE)
unemp_mn <- weighted.mean(pums_owners_v$unemp_rate_cbsa,   pums_owners_v$HHWT, na.rm = TRUE)
vac_mn   <- weighted.mean(pums_owners_v$vacancy_rate_cbsa, pums_owners_v$HHWT, na.rm = TRUE)
h_mn     <- weighted.mean(pums_owners_v$H_inc,             pums_owners_v$HHWT, na.rm = TRUE)
h_sd     <- sd(pums_owners_v$H_inc, na.rm = TRUE)

x_seq    <- seq(quantile(pums_owners_v$log_equiv_inc, 0.01, na.rm = TRUE),
                quantile(pums_owners_v$log_equiv_inc, 0.99, na.rm = TRUE),
                length.out = 100)
h_levels <- h_mn + c(-h_sd, 0, h_sd)
h_labels <- c("Low ($-$1 SD)", "Mean", "High ($+$1 SD)")

newdata_v <- expand.grid(
  log_equiv_inc     = x_seq,
  H_inc             = h_levels,
  race_eth          = factor("nhwhite", levels = levels(pums$race_eth)),
  college           = 0,
  AGE               = age_mn,
  gini_inc_cbsa     = gini_mn,
  H_bw              = Hbw_mn,
  unemp_rate_cbsa   = unemp_mn,
  vacancy_rate_cbsa = vac_mn
) |>
  mutate(H_label = factor(H_inc, levels = h_levels, labels = h_labels))

V_v_key <- v_vcov[c("log_equiv_inc", "H_inc", "log_equiv_inc:H_inc"),
                   c("log_equiv_inc", "H_inc", "log_equiv_inc:H_inc")]

newdata_v <- newdata_v |>
  mutate(
    y_hat = predict(v3, newdata = newdata_v),
    se    = mapply(function(x, h) {
              z <- c(x, h, x * h)
              sqrt(as.numeric(z %*% V_v_key %*% z))
            }, log_equiv_inc, H_inc),
    ylo   = y_hat - 1.96 * se,
    yhi   = y_hat + 1.96 * se
  )

ggplot(newdata_v, aes(x = exp(log_equiv_inc),
                      colour = H_label, fill = H_label, linetype = H_label)) +
  geom_ribbon(aes(ymin = ylo, ymax = yhi), alpha = 0.15, colour = NA) +
  geom_line(aes(y = y_hat), linewidth = 0.8) +
  scale_colour_manual(values = cb_palette[c(1, 3, 2)],
                      name   = TeX("Income Segregation $H^R$"),
                      labels = lapply(h_labels, TeX)) +
  scale_fill_manual(values   = cb_palette[c(1, 3, 2)],
                    name     = TeX("Income Segregation $H^R$"),
                    labels   = lapply(h_labels, TeX)) +
  scale_linetype_manual(values = lines[1:3],
                        name   = TeX("Income Segregation $H^R$"),
                        labels = lapply(h_labels, TeX)) +
  scale_x_continuous(labels = scales::label_dollar()) +
  labs(
    x = "Equivalised Household Income (1999 USD)",
    y = "Value-to-Income Ratio (predicted)"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
ggsave(paste0(results_dir, "burden_val_income_seg.pdf"), height = 12, width = 18, units = "cm")
```

### Imputed Cost-to-Income Ratio

```{r burden_cost_models}
pums_owners_c <- filter(pums, !is.na(cost_to_inc))

c1 <- feols(cost_to_inc ~ log_equiv_inc + race_eth + college + AGE + I(AGE^2) +
              gini_inc_cbsa + H_bw + unemp_rate_cbsa + vacancy_rate_cbsa | YEAR + CBSAFP,
            data = pums_owners_c, weights = ~HHWT, cluster = "CBSAFP")
c2 <- feols(cost_to_inc ~ log_equiv_inc + H_inc + race_eth + college + AGE + I(AGE^2) +
              gini_inc_cbsa + H_bw + unemp_rate_cbsa + vacancy_rate_cbsa | YEAR + CBSAFP,
            data = pums_owners_c, weights = ~HHWT, cluster = "CBSAFP")
c3 <- feols(cost_to_inc ~ log_equiv_inc * H_inc + race_eth + college + AGE + I(AGE^2) +
              gini_inc_cbsa + H_bw + unemp_rate_cbsa + vacancy_rate_cbsa | YEAR + CBSAFP,
            data = pums_owners_c, weights = ~HHWT, cluster = "CBSAFP")

modelsummary(
  list("(1)" = c1, "(2)" = c2, "(3)" = c3),
  estimate    = "{estimate}{stars}",
  statistic   = "({std.error})",
  stars       = c("+" = 0.1, "*" = .05, "**" = .01, "***" = .001),
  gof_map     = c("nobs", "r.squared", "FE: YEAR", "FE: CBSAFP"),
  output      = "html"
)
```

```{r burden_cost_plot}
c_coef <- coef(c3)
c_vcov <- vcov(c3)

age_mn   <- weighted.mean(pums_owners_c$AGE,               pums_owners_c$HHWT, na.rm = TRUE)
gini_mn  <- weighted.mean(pums_owners_c$gini_inc_cbsa,     pums_owners_c$HHWT, na.rm = TRUE)
Hbw_mn   <- weighted.mean(pums_owners_c$H_bw,              pums_owners_c$HHWT, na.rm = TRUE)
unemp_mn <- weighted.mean(pums_owners_c$unemp_rate_cbsa,   pums_owners_c$HHWT, na.rm = TRUE)
vac_mn   <- weighted.mean(pums_owners_c$vacancy_rate_cbsa, pums_owners_c$HHWT, na.rm = TRUE)
h_mn     <- weighted.mean(pums_owners_c$H_inc,             pums_owners_c$HHWT, na.rm = TRUE)
h_sd     <- sd(pums_owners_c$H_inc, na.rm = TRUE)

x_seq    <- seq(quantile(pums_owners_c$log_equiv_inc, 0.01, na.rm = TRUE),
                quantile(pums_owners_c$log_equiv_inc, 0.99, na.rm = TRUE),
                length.out = 100)
h_levels <- h_mn + c(-h_sd, 0, h_sd)
h_labels <- c("Low ($-$1 SD)", "Mean", "High ($+$1 SD)")

newdata_c <- expand.grid(
  log_equiv_inc     = x_seq,
  H_inc             = h_levels,
  race_eth          = factor("nhwhite", levels = levels(pums$race_eth)),
  college           = 0,
  AGE               = age_mn,
  gini_inc_cbsa     = gini_mn,
  H_bw              = Hbw_mn,
  unemp_rate_cbsa   = unemp_mn,
  vacancy_rate_cbsa = vac_mn
) |>
  mutate(H_label = factor(H_inc, levels = h_levels, labels = h_labels))

V_c_key <- c_vcov[c("log_equiv_inc", "H_inc", "log_equiv_inc:H_inc"),
                   c("log_equiv_inc", "H_inc", "log_equiv_inc:H_inc")]

newdata_c <- newdata_c |>
  mutate(
    y_hat = predict(c3, newdata = newdata_c),
    se    = mapply(function(x, h) {
              z <- c(x, h, x * h)
              sqrt(as.numeric(z %*% V_c_key %*% z))
            }, log_equiv_inc, H_inc),
    ylo   = y_hat - 1.96 * se,
    yhi   = y_hat + 1.96 * se
  )

ggplot(newdata_c, aes(x = exp(log_equiv_inc),
                      colour = H_label, fill = H_label, linetype = H_label)) +
  geom_ribbon(aes(ymin = ylo, ymax = yhi), alpha = 0.15, colour = NA) +
  geom_line(aes(y = y_hat), linewidth = 0.8) +
  scale_colour_manual(values = cb_palette[c(1, 3, 2)],
                      name   = TeX("Income Segregation $H^R$"),
                      labels = lapply(h_labels, TeX)) +
  scale_fill_manual(values   = cb_palette[c(1, 3, 2)],
                    name     = TeX("Income Segregation $H^R$"),
                    labels   = lapply(h_labels, TeX)) +
  scale_linetype_manual(values = lines[1:3],
                        name   = TeX("Income Segregation $H^R$"),
                        labels = lapply(h_labels, TeX)) +
  scale_x_continuous(labels = scales::label_dollar()) +
  labs(
    x = "Equivalised Household Income (1999 USD)",
    y = "Imputed Cost-to-Income Ratio (predicted)"
  ) +
  theme_bw() +
  theme(legend.position = "bottom")
ggsave(paste0(results_dir, "burden_cost_income_seg.pdf"), height = 12, width = 18, units = "cm")
```